The shape of kelp: environmental influences on kelp morphology

### 2018-09-06 chapter_2.Rmd
updated rmd file added

---
title: 'Chapter 2:The shape of kelp: environmental influences on kelp morphology'
author: "Ross Coppin"
date: "30 August 2018"
output:
  pdf_document: 
    highlight: zenburn
  word_document: default
---
## Introduction

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], nevertheless form the basis of one of the most productive ecosystems globally [@Mann1973]. Kelps generally have a dependence on cool, temperate and arctic seawater temperatures [@Santelices2007; @Bolton2010], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Outside of these latitudes, they are also found in cool, deep water towards the tropics [@Graham2007]: this deeper tropical water, due to its water clarity, also allows for a minimum photon flux rate of 50 ??? and a nitrate concentration of ??? [@Luning1990; @Dayton1999; @Zimmerman1985], the two other environmental variables that support kelp productivity. Although kelps have a low taxonomic diversity, their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude other turf and subcanopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Dunton1987; @Duggins1989; @Steneck2002], each depending on a wide suite of ecological services provided by the kelp habitat [@Gaines1987; @Bologna1993; @Levin1994; @Anderson1997].

@@

The biogeographic distribution of kelp is limited by seawater temperature [@Bolton2010], where increasing temperature gradients reduce kelp distribution. Due to this limiting factor, the two main species of kelps in southern African waters, _Ecklonia maxima_ and _Laminaria pallida_, are distributed along a section of the south coast from De Hoop, extending west around the Cape Peninsula, and thriving north into Namibia [@Molloy1990; @Stegenga1997]. This distribution follows a temperate gradient, where sea temperatures increase as one moves south from Namibia, around Cape Point and towards De Hoop. Although the two species occur together for the majority of the coast, their basic morphologies and resource needs vary to a degree. The larger species, _Ecklonia maxima_, is distributed from Lüderitz to Cape Agulhas [@Bolton1985; @Probyn1985; @Bolton1987; @Bolton2012]. Characterised by a large distal swollen bulb filled with gas, and smooth fronds, this species grows to approximately 10 meters [@Bolton1987]. There was, however, a 17-meter specimen collected in 2015 off Cape Point (Smit, unpubl. data).This species of kelp not only dominate the biomass of the South African nearshore, but plays an important ecological role [@Bustamante1996]. The estimated productivity of _Ecklonia maxima_ within South Africa varies between 350 and 1500g Cm-2yr-1 [@Mann1982]. Across the majority of the coastline, _Laminaria pallida_ remains a subsurface kelp, dominating the kelp biomass at depths greater than 10 meters [@Field1980; @Bolton1987; @Molloy1996]. This species is distributed from Danger Point, east of the Cape Peninsula, to Rocky Point in northern Namibia, and reaches depths of greater than 20 meters [@Field1980; @Molloy1990; @Molloy1996; @Stegenga1997]. Towards the north along the west coast, from around Hondeklipbaai, _Laminaria pallida_ replaces _Ecklonia maxima_ as the dominant kelp species [@Velimirov1977; @Stegenga1997] and it also occupies increasingly shallower subtidal regions. The northern populations also exhibit an increase in stipe hollowness, compared to the solid stipe morphs in the species’ southern distributions [@Molloy1996]. This variation in morphology was thought to represent two distinct species, with the northern populations formerly described as _Laminaria schinzii_ Foslie [@Molloy1996]. Genetic work has subsequently shown that the two morphs are in fact the same species [@Rothman2017]. In southern African waters, the primary production of _Laminaria pallida_ is between 120 and 1900g C m2yr1, similar to that of _Ecklonia maxima_ [@Mann1982]. Primary production is not the only pathway.

@@

In order to compare abiotic parameters for sites around the coast, large historical databases for both temperature and wave energy were accessed. Shallow water temperatures were sourced from the South African Coastal Temperature Network (SACTN) website (https://github.com/ajsmit/SACTN). Seven different organisations within South Africa contribute to the SACTN, where _in situ_ temperature measurements are made around the South African coast using either hand-held thermometers or digital temperature recorders positioned underwater. The mean duration of the 135 daily time series is 19.7 years, and these _in situ_ data are preferred over satellite SST, which have shown to exhibit large biases [@Smit2013]. Linear interpolated SST were calculated for sites where _in situ_ recorders were absent. Wave energy data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs [@Theron2014]. These data are first forecasted using NOAA Wave Watch III (WWIII), with National Centers for Environmental Predictions (NCEP) product as the numerical input [@benassai2006implementation]. Hindcast data from WWIII span from 1994-2013 at a 3-hour resolution. The data are then used to model short –crested waves generated by the wind into the coastal environment, using Simulating Waves in the Nearshore [SWAN; @booij1997swan]. SWAN allows one to extract wave parameters from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modelled, at both 7 meter and 15 meter contours. A 200 meter resolution was used as False Bay was nested within a larger grid area for the research from where the data were sourced.  For Table Bay and east of Cape Hangklip the resolution 500 meters at 7 meters and 15 meters. For this study the 7 meter contours were used.

@@

# Laminaria
lam_data <- read_csv("data/Laminaria_morphometrics.csv") %>%
  mutate(site = replace(site, site == "Buffels Bay", "Buffelsbaai")) %>%
  # NB: Even though I am changing the name, there will still be no lon/lat values for this site
  # We need to decide how we want to handle this. Give it the same lon/lat as Buffelsbaai?
  mutate(site = replace(site, site == "Buffels Bay South", "Buffelsbaai South")) %>%
  # NB: This may also not be an appropriate change
  mutate(site = replace(site, site == "Bordjiestif North", "Bordjies Deep")) %>%
  # Not using Rocky Bank in this study
  filter(site != "Rocky Bank")

@@

# Calculate the necessary stats
# Monthly clims
wave_feb_aug <- wave_model %>% 
  mutate(date = lubridate::month(date, label = TRUE)) %>% 
  filter(site!= "FB25") %>% # FB_25 has issues with 'hs' and 'tp'
  group_by(site, date) %>% 
  summarise_all(funs(mean = mean, sd = sd), na.rm = T) %>%
  filter(date %in% c("Feb", "Aug")) %>% 
  ungroup()

@@

max <- ggplot(data = maxTemp, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = maxTemp, aes(col = value), size = 2) +
  scale_color_gradient("Max temp (°)", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(season), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

min <- ggplot(data = minTemp, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = minTemp, aes(col = value), size = 2) +
  scale_color_gradient("Min temp (°)", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(season), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

mean <- ggplot(data = meanTemp, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = meanTemp, aes(col = value), size = 2) +
  scale_color_gradient("Mean temp (°)", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(season), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

sd <- ggplot(data = sdTemp, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = sdTemp, aes(col = value), size = 2) +
  scale_color_gradient("SD temp (°)", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(season), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

range <- ggplot(data = rangeTemp, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = rangeTemp, aes(col = value), size = 2) +
  scale_color_gradient("Range temp (°)", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(season), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

@@

# meanDir
medianDir1 <- ggplot(data = medianDir, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = medianDir, aes(col = value), size = 2) +
  scale_color_gradient("medianDir", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), scales = "free", ncol = 3) + 
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# meanDirw 
medianDirw1 <- ggplot(data = medianDirw , aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = medianDirw , aes(col = value), size = 2) +
  scale_color_gradient("medianDirw ", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

@@

meanHs1 <- ggplot(data = meanHs, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = meanHs, aes(col = value), size = 2) +
  scale_color_gradient("meanHs", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# sdHs
sdHs1 <- ggplot(data = sdHs, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = sdHs, aes(col = value), size = 2) +
  scale_color_gradient("sdHs", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# meanSpw
meanSpw1 <- ggplot(data = meanSpw, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = meanSpw, aes(col = value), size = 2) +
  scale_color_gradient("meanSpw", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# sdSpw
sdSpw1 <- ggplot(data = sdSpw, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = sdSpw, aes(col = value), size = 2) +
  scale_color_gradient("sdSpw", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# meanTp
meanTp1 <- ggplot(data = meanTp, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = meanTp, aes(col = value), size = 2) +
  scale_color_gradient("meanTp", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# sdTp
sdTp1 <- ggplot(data = sdTp, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = sdTp, aes(col = value), size = 2) +
  scale_color_gradient("sdTp", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), scales = "free", ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())
```

@@

Not only do the temperature parameters display seasonal fluctuations, but we see similar observations for wave parameters. The west side of Cape Point exhibit increased measurements of median direction, SD and mean significant wave height compared to within False Bay (Fig. 4). Sites within False Bay however display increased standard deviations of wave period. Median wind direction is northerly in winter (Fig. 5), and turns southerly in summer.

@@

#### _Laminaria pallida_
Lamina length for laminaria pallida showed no geographical pattern moving from west to east. Kommetjie, Olifantsbos and Batsata Rock showed great variability in lamina length, and both Buffels Bay and Betty’s Bay were visually different to Miller’s Point and Roman Rock, when comparing the summary data represented by the boxplots (Fig.6). Lamina thickness showed great variation across sites, with Baboon Rock, Miller’s Point, A-Frame and Roman rock displaying large lamina thickness, significantly different to the rest of the sites. Lamina weight was observed to vary for Baboon Rock and Betty’s Bay. Neighbouring sites A-Frame and Roman Rock also showed visual difference when comparing boxplot summary statistics. An increase in the number of digits was observed as one moved from Cape Point north along the western side of False Bay. This ceased at Batstata Rock, which exhibited significantly less digits compared to the previous site, Bordjiestif North. Stipe diameter showed some geographical grouping, with west of Cape Point sites exhibiting larger stipe diameters compared to False Bay sites.

Stipe diameter decreased as one rounded the point into False Bay, where a sudden, significant difference was seen between Bordjiestif North and Batsata Rock. Greater variation of stipe length was observed for sites outside of False Bay (Kommetjie, Olifantsbos and Betty’s Bay). Baboon Rock, Miller’s Point, A-Frame and Roman Rock were again grouped together exhibiting the lowest stipe lengths, that were all significantly different to sites found west of Cape Point. Stipe mass displays similar patterns to stipe length, with larger stipe lengths west of Cape Point compared to within False Bay. The thallus mass of _Laminaria pallida_ was observed to be greater for sites around Cape Point, with an observed difference between Batsata Rock and Bordjiestif North, similar to the number of digit patterns. Larger total lengths were observed around Cape Point and Betty’s Bay, with the smallest total lengths at sites that exhibited the greatest lamina thicknesses (Baboon Rock, Miller’s Point, A-Frame and Roman Rock).

@@

# Order site names by location west to east
positionsLam <- c("Betty's Bay", "Roman Rock","A-Frame", "Miller's Point", "Baboon Rock", 
                  "Batsata Rock", "Bordjiestif North", "Buffels Bay South", 
                  "Buffels Bay", "Olifantsbos", "Kommetjie", "Sea Point")

>>>>

---
title: "The shape of kelp: environmental influences on kelp morphology"
output:
  word_document:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    toc: no
  html_notebook:
    fig_caption: yes
    toc: yes
  pdf_document:
    toc: yes
  bookdown::html_document2:
    fig_caption: yes
    fig_retina: 2
    fig_width: 7
    highlight: espresso
    theme: sandstone
    toc_float: no
fontsize: 12pt
geometry: margin=1in
language: Australian
mainfont: PT Serif
monofont: PT Mono
csl: LaTeX/ecography.csl
papersize: A4
sansfont: PT Sans
tables: yes
bibliography: LaTex/mycollection.bib
---

\newpage
## Introduction

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], nevertheless form the basis of one of the most productive ecosystems globally [@Mann1973]. Kelps generally have a dependence on cool, temperate and arctic seawater temperatures [@Santelices2007; @Bolton2010], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Outside of these latitudes, they are also found in cool, deep water towards the tropics [@Graham2007]: this deeper tropical water, due to its water clarity, also allows for suitable photon flux and a nitrate concentration [@Luning1990; @Dayton1999; @Zimmerman1984], the two other environmental variables that support kelp productivity. Although kelps have a low taxonomic diversity, their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude other turf and subcanopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Dunton1987; @Duggins1989; @Steneck2002], each depending on a wide suite of ecological services provided by the kelp habitat [@Gaines1987; @Bologna1993; @Levin1994; @Anderson1997].

@@

The biogeographic distribution of kelp is limited by seawater temperature [@Bolton2010], where increasing temperature gradients reduce kelp distribution. Due to this limiting factor, the two main species of kelps in southern African waters, _Ecklonia maxima_ and _Laminaria pallida_, are distributed along a section of the south coast from De Hoop, extending west around the Cape Peninsula, and thriving north into Namibia [@Molloy1990; @Stegenga1997]. This distribution follows a temperate gradient, where sea temperatures increase as one moves south from Namibia, around Cape Point and towards De Hoop. Although the two species occur together for the majority of the coast, their basic morphologies and resource needs vary to a degree. The larger species, _Ecklonia maxima_, is distributed from Lüderitz to Cape Agulhas [@Bolton1985a; @Probyn1985; @Bolton1987; @Bolton2012]. Characterised by a large distal swollen bulb filled with gas, and smooth fronds, this species grows to approximately 10 meters [@Bolton1987]. There was, however, a 17-meter specimen collected in 2015 off Cape Point (Smit, unpubl. data).This species of kelp not only dominate the biomass of the South African nearshore, but plays an important ecological role [@Bustamante1996]. The estimated productivity of _Ecklonia maxima_ within South Africa varies between 350 and 1500g Cm-2yr-1 [@Mann1982]. Across the majority of the coastline, _Laminaria pallida_ remains a subsurface kelp, dominating the kelp biomass at depths greater than 10 meters [@Field1980b; @Bolton1987; @Molloy1996]. This species is distributed from Danger Point, east of the Cape Peninsula, to Rocky Point in northern Namibia, and reaches depths of greater than 20 meters [@Field1980b; @Molloy1990; @Molloy1996; @Stegenga1997]. Towards the north along the west coast, from around Hondeklipbaai, _Laminaria pallida_ replaces _Ecklonia maxima_ as the dominant kelp species [@Velimirov1977; @Stegenga1997] and it also occupies increasingly shallower subtidal regions. The northern populations also exhibit an increase in stipe hollowness, compared to the solid stipe morphs in the species’ southern distributions [@Molloy1996]. This variation in morphology was thought to represent two distinct species, with the northern populations formerly described as _Laminaria schinzii_ Foslie [@Molloy1996]. Genetic work has subsequently shown that the two morphs are in fact the same species [@Rothman2017]. In southern African waters, the primary production of _Laminaria pallida_ is between 120 and 1900g C m2yr1, similar to that of _Ecklonia maxima_ [@Mann1982]. Primary production is not the only pathway.

@@

In order to compare abiotic parameters for sites around the coast, large historical databases for both temperature and wave energy were accessed. Shallow water temperatures were sourced from the South African Coastal Temperature Network (SACTN) website (https://github.com/ajsmit/SACTN). Seven different organisations within South Africa contribute to the SACTN, where _in situ_ temperature measurements are made around the South African coast using either hand-held thermometers or digital temperature recorders positioned underwater. The mean duration of the 135 daily time series is 19.7 years, and these _in situ_ data are preferred over satellite SST, which have shown to exhibit large biases [@Smit2013]. Linear interpolated SST were calculated for sites where _in situ_ recorders were absent. Wave energy data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs [@Theron2014]. These data are first forecasted using NOAA Wave Watch III (WWIII), with National Centers for Environmental Predictions (NCEP) product as the numerical input [@benassai2006]. Hindcast data from WWIII span from 1994-2013 at a 3-hour resolution. The data are then used to model short –crested waves generated by the wind into the coastal environment, using Simulating Waves in the Nearshore [SWAN; @booij1996]. SWAN allows one to extract wave parameters from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modelled, at both 7 meter and 15 meter contours. A 200 meter resolution was used as False Bay was nested within a larger grid area for the research from where the data were sourced.  For Table Bay and east of Cape Hangklip the resolution 500 meters at 7 meters and 15 meters. For this study the 7 meter contours were used.

@@

# Laminaria
lam_data <- read_csv("data/Laminaria_morphometrics.csv") %>%
  mutate(site = replace(site, site == "Buffels Bay", "Buffelsbaai")) %>%
  # NB: Even though I am changing the name, there will still be no lon/lat values for this site
  # We need to decide how we want to handle this. Give it the same lon/lat as Buffelsbaai?
  mutate(site = replace(site, site == "Buffels Bay South", "Buffelsbaai South")) %>%
  # NB: This may also not be an appropriate change
  mutate(site = replace(site, site == "Bordjies reef North", "Bordjies Deep")) %>%
  # Not using Rocky Bank in this study
  filter(site != "Rocky Bank")

@@

# Calculate the necessary stats
# Monthly clims
wave_feb_aug <- wave_model %>% 
  mutate(date = lubridate::month(date, label = TRUE)) %>% 
  filter(site != "FB25") %>% # FB_25 has issues with 'hs' and 'tp'
  group_by(site, date) %>% 
  summarise_all(funs(mean = mean, sd = sd), na.rm = T) %>%
  filter(date %in% c("Feb", "Aug")) %>% 
  ungroup()

@@

max <- ggplot(data = maxTemp, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = maxTemp, aes(col = value), size = 2) +
  scale_color_gradient("Max temp (°)", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(season), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

min <- ggplot(data = minTemp, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = minTemp, aes(col = value), size = 2) +
  scale_color_gradient("Min temp (°)", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(season), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

mean <- ggplot(data = meanTemp, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = meanTemp, aes(col = value), size = 2) +
  scale_color_gradient("Mean temp (°)", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(season), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

sd <- ggplot(data = sdTemp, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = sdTemp, aes(col = value), size = 2) +
  scale_color_gradient("SD temp (°)", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(season), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

range <- ggplot(data = rangeTemp, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = rangeTemp, aes(col = value), size = 2) +
  scale_color_gradient("Range temp (°)", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(season), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

@@

# meanDir
medianDir1 <- ggplot(data = medianDir, aes(x = lon, y = lat)) +
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = medianDir, aes(col = value), size = 2) +
  scale_color_gradient("medianDir", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), ncol = 3) + 
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# meanDirw 
medianDirw1 <- ggplot(data = medianDirw , aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = medianDirw , aes(col = value), size = 2) +
  scale_color_gradient("medianDirw ", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

@@

meanHs1 <- ggplot(data = meanHs, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = meanHs, aes(col = value), size = 2) +
  scale_color_gradient("meanHs", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# sdHs
sdHs1 <- ggplot(data = sdHs, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = sdHs, aes(col = value), size = 2) +
  scale_color_gradient("sdHs", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# meanSpw
meanSpw1 <- ggplot(data = meanSpw, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = meanSpw, aes(col = value), size = 2) +
  scale_color_gradient("meanSpw", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# sdSpw
sdSpw1 <- ggplot(data = sdSpw, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = sdSpw, aes(col = value), size = 2) +
  scale_color_gradient("sdSpw", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# meanTp
meanTp1 <- ggplot(data = meanTp, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = meanTp, aes(col = value), size = 2) +
  scale_color_gradient("meanTp", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

# sdTp
sdTp1 <- ggplot(data = sdTp, aes(x = lon, y = lat)) + 
  geom_polygon(data = south_africa_coast, aes(group = group), fill = "grey70") +
  geom_path(data = sa_provinces, aes(group = group)) +
  geom_point(data = sdTp, aes(col = value), size = 2) +
  scale_color_gradient("sdTp", low = "blue", high = "red") +
  coord_fixed(xlim = c(17.85, 19), ylim = c(-32.75, -34.4)) +
  facet_wrap(~ season, labeller = as_labeller(c(season)), ncol = 3) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())
```

@@

Not only do the temperature parameters display seasonal fluctuations, but we see similar observations for wave parameters. The west side of Cape Point exhibit increased median direction, SD and mean significant wave height compared to within False Bay (Fig. 4). Sites within False Bay however display increased standard deviations of wave period. Median wind direction is northerly in winter (Fig. 5), and turns southerly in summer.

@@

#### _Laminaria pallida_
Lamina length for laminaria pallida showed no geographical pattern moving from west to east. Kommetjie, Olifantsbos and Batsata Rock showed great variability in lamina length, and both Buffels Bay and Betty’s Bay were visually different to Miller’s Point and Roman Rock, when comparing summary data (Fig.6). Lamina thickness showed great variation across sites, with Baboon Rock, Miller’s Point, A-Frame and Roman rock displaying large lamina thickness, significantly different to the rest of the sites. Lamina weight was observed to vary for Baboon Rock and Betty’s Bay. Neighbouring sites A-Frame and Roman Rock also showed visual difference when comparing boxplot summary statistics. An increase in the number of digits was observed as one moved from Cape Point north along the western side of False Bay. This ceased at Batstata Rock, which exhibited significantly less digits compared to the previous site, Bordjies reef North. Stipe diameter showed some geographical grouping, with west of Cape Point sites exhibiting larger stipe diameters compared to False Bay sites.

Stipe diameter decreased as one rounded the point into False Bay, where a sudden, significant difference was seen between Bordjies reef North and Batsata Rock. Greater variation of stipe length was observed for sites outside of False Bay (Kommetjie, Olifantsbos and Betty’s Bay). Baboon Rock, Miller’s Point, A-Frame and Roman Rock were again grouped together exhibiting the lowest stipe lengths, that were all significantly different to sites found west of Cape Point. Stipe mass displays similar patterns to stipe length, with larger stipe lengths west of Cape Point compared to within False Bay. The thallus mass of _Laminaria pallida_ was observed to be greater for sites around Cape Point, with an observed difference between Batsata Rock and Bordjies reef North, similar to the number of digit patterns. Larger total lengths were observed around Cape Point and Betty’s Bay, with the smallest total lengths at sites that exhibited the greatest lamina thicknesses (Baboon Rock, Miller’s Point, A-Frame and Roman Rock).

@@

# Order site names by location west to east
positionsLam <- c("Betty's Bay", "Roman Rock","A-Frame", "Miller's Point", "Baboon Rock", 
                  "Batsata Rock", "Bordjies reef North", "Buffels Bay South", 
                  "Buffels Bay", "Olifantsbos", "Kommetjie", "Sea Point")
